 

### 2021-02-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

### Light environments
**Light conditions, specifically changes in proportion incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory. Shaded regions experience lower values of PAR with different light quality [in ranges of blue, orange and red]. The light gradient is more pronounced under broadleaf forests than needleleaf forests [@lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b]
(Fig. `r fig_NEON_vertical`).

@@

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
Across the **#** forested NEON sites, windspeeds universally increased with height, with daily maximum wind speeds averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom. 
Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].

@@

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. In northeastern US mesic temperate forests, tree-ring analyis [__time-scale] showed that the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]

@@

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; Rollinson et al., in press].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).
It is also possible that competition is accelerated under warmer temperatures.

>>>>

### Light environments
**Light conditions, specifically changes in proportion incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory. Shaded regions experience lower values of PAR with different light quality [in ranges of blue, orange and red]. 
The light gradient is more pronounced under broadleaf forests than needleleaf forests [Fig. `r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b].

@@

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [**REFS**].**
Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom. 
Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].

@@

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].

@@

In northeastern US mesic temperate forests, tree-ring analyis [__time-scale] showed that the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).
It is also possible that competition is accelerated under warmer temperatures.
